---
title: "austraits"
author: "Fonti Kar, Daniel Falster"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
  toc: yes
vignette: >
  %\VignetteIndexEntry{austraits}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  comment = "#>"
)

library(dplyr)
library(stringr)
library(austraits)
```

The `austraits` package is designed to aid users to access data from AusTraits, a curated plant trait database for the Australian flora. This package contains several core functions to wrangle and visualise data. Below we included a tutorial to illustrate how to use these functions in greater detail. 

## Getting started
`austraits` is still under development. To install the current version from github 

```{r setup, eval = FALSE}
install.packages("remotes")
remotes::install_github("traitecoevo/austraits", build_vignettes = TRUE)

# Load the austraits package
library(austraits)
```

### Meta-data for austraits

Definitions for the various tables and traits are found under Articles.

### Load the data

```{r, load_data, eval = FALSE}
austraits <- load_austraits()
```

Note the `austraits` object is a list with a bunch of elements:

```{r}
names(austraits)
```

## Extracting data

AusTraits contains `r length(unique(austraits$traits$trait_name))` plant traits. In most cases, users would like to extract a subset of the dataset for their own research purposes. `extract_dataset`extracts a particular study, whereas `extract_trait`subsets by certain traits. Note that the other tables and elements of the AusTraits data are extracted too, not just the main trait table. See `?extract_dataset` and `?extract_trait` for more details

### Extracting by study

Filtering one particular study and assigning it to an object

```{r, extract_study}
data_zanne_09 <- extract_dataset(austraits, "Zanne_2009")

data_zanne_09$traits
```

Filtering multiple studies by same lead author (e.g. Falster) and assigning it to an object

```{r, extract_studies}
data_falster_studies <- extract_dataset(austraits, 
                                        dataset_id = c("Falster_2003","Falster_2005_1"))

data_falster_studies$traits
```

Identify all studies with an id that includes "Falster"

```{r, extract_Falster_studies}
( dataset_ids <- austraits$methods$dataset_id %>% unique() %>% subset(., grepl("Falster",.))) 

data_falster_studies <- extract_dataset(austraits, dataset_ids)

data_falster_studies$traits
```

### Extracting by trait

Filtering one particular trait and assigning it to an object

```{r, extract_trait}
data_wood_dens <- extract_trait(austraits, "wood_density")

data_wood_dens$traits
```

Filtering all traits with leaf in the name
```{r}
( leaf_traits <- austraits$traits$trait_name %>% unique() %>% subset(., grepl("leaf",.))) 

data_leaf <- extract_trait(austraits, leaf_traits) # Extract from austraits

data_leaf$traits
```


## Join data from other tables and elements

Once users have extracted the data they want, they may want to merge in other study details into the main traits dataframe for their analyses. For example, users may require taxonomic information for a phylogenetic analysis. This is where the `join_` functions come in. There are five `join_` functions in total, each designed to append specific information from other tables and elements in the `austraits` object. Their suffixes refer to the type of information that is joined, e.g. `join_taxonomy` appends taxonomic information to the `traits` dataframe. See `?join_all` for more details. 

```{r, join_}
# Join taxonomic information 
(data_leaf %>% join_taxonomy)$traits

# Join methodological information 
(data_leaf %>% join_methods)$traits

# Join site based information 
(data_leaf %>% join_sites)$traits

# Join context information 
(data_leaf %>% join_contexts)$traits

# Alternatively users can join *all* information 
(data_leaf %>% join_all)$traits
```

## Visualising data by site

`plot_site_locations` graphically summarises where trait data was collected from and how much data is available. The legend refers to the number of neighbouring points, the more warmer the colour, the more data is available. This function only works for studies that are geo-referenced. Users must first use `join_sites` to append  latitude and longtitude information into the trait dataframe before plotting

```{r, site_plot}
data_wood_dens <- data_wood_dens %>% join_sites
data_wood_dens$traits %>% plot_site_locations()
```

## Visualising data distribution and variance

`plot_trait_distribution` creates histograms and [beeswarm plots](https://github.com/eclarke/ggbeeswarm) for specific traits to help users visualise the variance of the data. Users can specific whether to create seperate beeswarm plots at the level of taxonomic family or for each `dataset_id`

```{r, beeswarm}
austraits %>% plot_trait_distribution_beeswarm("wood_density", "family")
austraits %>% plot_trait_distribution_beeswarm("wood_density", "dataset_id")
```

### Pivotting from long to wide  format

The table of traits in AusTraits comes in long format, where data for all traits are in two columns called `trait_name` and `value`. You can convert to wide format, where each trait is in a separate column, using the function `trait_pivot_wider`.

```{r, pivot}
data_wide <- data_falster_studies$traits %>% trait_pivot_wider()
```

The returned object is a list with several parts:
```{r}
data_wide %>% names()
```

Looking at the `value` element, it now has columns for each trait:

```{r}
data_wide$value %>% names()
data_wide$value
```

