---
title: "austraits"
author: "Fonti Kar, Daniel Falster"
date: "2-Jun-2021"
output: 
  rmarkdown::html_vignette:
  toc: yes
vignette: >
  %\VignetteIndexEntry{austraits}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  comment = "#>"
)
```

The `austraits` package is designed to aid users to access data from AusTraits, a curated plant trait database for the Australian flora. This package contains several core functions to wrangle data for their purposes. We have also included a tutorial below to illustrate how to use these functions in greater detail. 

### Getting started
`austraits` is still under development. To install the current version from github 
```{r setup, eval = FALSE}
install.packages("remotes")
remotes::install_github("traitecoevo/austraits", build_vignettes = TRUE)

#Load the austraits package
library(austraits)
```

### Filter data
```{r}

```

### Extract data
```{r}

```

### Join other information
```{r}

```

### General data visualisations
```{r}

```

