---
title: "austraits"
author: "Fonti Kar, Daniel Falster"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
  toc: yes
vignette: >
  %\VignetteIndexEntry{austraits}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  comment = "#>"
)

library(dplyr)
library(stringr)
library(austraits)

austraits <- load_austraits(version = "3.0.2", path = "ignore/data/austraits")
```

The `austraits` package is designed to aid users to access data from [AusTraits](https://zenodo.org/record/5112001), a curated plant trait database for the Australian flora. This package contains several core functions to wrangle and visualise data. Below we included a tutorial to illustrate how to use these functions in greater detail. 

## Getting started
`austraits` is still under development. To install the current version from GitHub 

```{r setup, eval = FALSE}
#install.packages("remotes")
remotes::install_github("traitecoevo/austraits", build_vignettes = TRUE)

# Load the austraits package
library(austraits)
```

### Loading AusTraits database

By default, `load_austraits` will download AusTraits to a specified path e.g. `data/austraits` and will reload it from this location in the future. You can set `update = TRUE` so the austrait versions are downloaded fresh from [Zenodo](https://zenodo.org/record/3568429). Note that `load_austraits` will happily accept a DOI of a particular version.

```{r, load_data, eval = FALSE}
austraits <- load_austraits(version = "3.0.2", path = "data/austraits")
```

You can check out different versions and their associated DOI of the AusTraits by using: 

```{r, eval = FALSE}
get_versions(path = "data/austraits")
```

```{r, echo=FALSE}
get_versions(path = "ignore/data/austraits")
```

The AusTraits object is a very *long* list with various of elements. If you are not familiar with working with lists in R, we recommend having a quick look at this [tutorial](https://www.tutorialspoint.com/r/r_lists.htm). To learn more about the structure of `austraits`, check out the [structure of the database](https://traitecoevo.github.io/austraits/articles/structure.html). 

```{r}
class(austraits)

# The names of the different lists in `austraits`
names(austraits)

# The structure of the `traits` table only
str(austraits$traits)
```

## Descriptive summaries of traits and taxa

AusTraits contains `r length(unique(austraits$traits$trait_name))` plant traits. Check out [definitions of the traits](https://traitecoevo.github.io/austraits/articles/dictionary.html) to learn more about how each trait is defined. 

Have a look at what trait or taxa we have with: 

```{r}
print_austraits(austraits, "trait_name") %>% head()

print_austraits(austraits, "family") %>% head()

print_austraits(austraits, "genus") %>% head()
```

Interested in a specific trait? Try `lookup_trait`

```{r}
lookup_trait(austraits, "leaf") %>% head()
```

## Extracting data

In most cases, users would like to extract a subset of `austraits` for their own research purposes.`extract_dataset` subsets a particular study, whereas `extract_trait`subsets by certain traits. Note that the other tables and elements of the AusTraits data are extracted too, not just the main trait table. See `?extract_dataset` and `?extract_trait` for more details

### Extracting by study

Filtering **one particular study** and assigning it to an object

```{r, extract_study}
data_zanne_09 <- extract_dataset(austraits, "Zanne_2009")

data_zanne_09$traits %>% head()
```

Filtering **multiple studies by two different lead authors**  and assigning it to an object

```{r, extract_studies}
data_falster_studies <- extract_dataset(austraits, 
                                        dataset_id = c("Falster_2003","Zanne_2009"))
 
data_falster_studies$traits %>% head()
```

Filtering **multiple studies by same lead author** (e.g. Falster) and assigning it to an object.  

```{r, extract_Falster_studies}
# First, we need to identify all studies with an id that includes "Falster"

( dataset_ids <- austraits$methods$dataset_id %>% unique() %>% subset(., grepl("Falster",.))) 

# Then we extract
data_falster_studies <- extract_dataset(austraits, dataset_ids)

data_falster_studies$traits %>% head()
```

### Extracting by taxonomic level

```{r extract_taxa}
# By family 
proteaceae <- extract_taxa(austraits, family = "Proteaceae")
# Checking that only taxa in Proteaceae have been extracted
proteaceae$taxa$family %>% unique()

# By genus 
acacia <- extract_taxa(austraits, genus = "Acacia")
# Checking that only taxa in Acacia have been extracted
acacia$traits$taxon_name %>% unique() %>% head()
```

### Extracting by trait

Filtering **one particular trait** and assigning it to an object

```{r, extract_trait}
data_wood_dens <- extract_trait(austraits, "wood_density")

head(data_wood_dens$traits)
```

Combining `lookup_trait` with `extract_trait` to obtain **all traits with 'leaf' in the trait name** and assigning it to an object. Note we use the `.` notation to pass on the `lookup_trait` results to `extract_trait`

```{r}
data_leaf <- lookup_trait(austraits, "leaf") %>% extract_trait(austraits, .) 

head(data_leaf$traits)
```

## Join data from other tables and elements

Once users have extracted the data they want, they may want to merge in other study details into the main `traits` dataframe for their analyses. For example, users may require taxonomic information for a phylogenetic analysis. This is where the `join_` functions come in. 

There are five `join_` functions in total, each designed to append specific information from other tables and elements in the `austraits` object. Their suffixes refer to the type of information that is joined, e.g. `join_taxonomy` appends taxonomic information to the `traits` dataframe. See `?join_all` for more details. 

```{r, join_}
# Join taxonomic information 
(data_leaf %>% join_taxonomy)$traits %>% head()

# Join methodological information 
(data_leaf %>% join_methods)$traits %>% head()

# Join site based information 
(data_leaf %>% join_sites)$traits %>% head()

# Join context information 
(data_leaf %>% join_contexts)$traits %>% head()

# Alternatively users can join *all* information 
(data_leaf %>% join_all)$traits %>% head()
```

## Visualising data by site

`plot_site_locations` graphically summarises where trait data was collected from and how much data is available. The legend refers to the number of neighbouring points, the more warmer the colour, the more data is available. This function only works for studies that are geo-referenced. Users must first use `join_sites` to append latitude and longitude information into the trait dataframe before plotting

```{r, site_plot, fig.align = "center", fig.width=5, fig.height=5}
data_wood_dens <- data_wood_dens %>% join_sites()
plot_site_locations(data_wood_dens$traits)
```

## Visualising data distribution and variance

`plot_trait_distribution` creates histograms and [beeswarm plots](https://github.com/eclarke/ggbeeswarm) for specific traits to help users visualise the variance of the data. Users can specific whether to create seperate beeswarm plots at the level of taxonomic family or for each `dataset_id`

```{r, beeswarm, fig.align = "center", fig.width=6, fig.height=7}
austraits %>% plot_trait_distribution_beeswarm("wood_density", "family")
austraits %>% plot_trait_distribution_beeswarm("wood_density", "dataset_id")
```

### Pivotting from long to wide format

The table of traits in AusTraits comes in long format, where data for all traits are in two columns called `trait_name` and `value`. You can convert to wide format, where each trait is in a separate column, using the function `trait_pivot_wider`. This will produce 5 separate lists (`value`, `unit`, `date`, `value_type` and `replicates`). The trait values are stored in the `value` list.

Note that in some cases, there are *multiple observations* for some traits, e.g. huber values. This can prevent pivoting from long format to wide. In this circumstances, we suggest using `bind_trait_values` which joins multiple observations together into a string.

```{r, pivot}
data_wide_bound <- data_falster_studies$traits %>%
  bind_trait_values() %>% # Joining multiple obs with `--`
  trait_pivot_wider() #Pivot wide

data_wide_bound$value # The trait values table

# Check out the 'bounded' trait values
data_wide_bound$value$huber_value[!is.na(data_wide_bound$value$huber_value)] %>% head()
```

If you would like to revert the bounded trait values, you have to `trait_pivot_longer` first then: 

```{r}
data_wide_bound %>% 
  trait_pivot_longer() %>% 
  separate_trait_values(., austraits$definitions)
```

Alternatively, if you don't want to bind the trait values, you can give `summarise_trait_means` a go, which computers means with the multiple observations. Note that this method condenses the traits table and you won't be able to revert it back unless you `load_austraits` again

```{r, pivot_summarised}
data_wide_summarised <- data_falster_studies$traits %>%
  summarise_trait_means() %>% 
  trait_pivot_wider()

data_wide_summarised$value %>% head()
```

